dataaimsr: An R Client for the Australian Institute of Marine Science Data Platform API which provides easy access to AIMS Data Platform

### 2021-03-18 paper/paper.Rmd
updated paper text and map

---
title: 'dataaimsr: An R Client for the Australian Institute of Marine Science Data Platform API which provides easy access to AIMS Data Platform'
tags:
  - R
  - sea surface temperature
  - weather stations
  - long-term environmental monitoring
  - Australia
  - API
authors:
  - name: Diego R. Barneche
    orcid: 0000-0002-4568-2362
    affiliation: "1, 2"
  - name: Greg Coleman
    affiliation: "3"
  - name: Jeffrey L. Sheehan
    affiliation: "3"
  - name: Duncan Fermor
    affiliation: "3"
  - name: Murray Logan
    affiliation: "3"
  - name: Mark Rehbein
    affiliation: "3"
affiliations:
 - name: Australian Institute of Marine Science, Crawley, WA 6009, Australia
   index: 1
 - name: The Indian Ocean Marine Research Centre, The University of Western Australia, Crawley, WA 6009, Australia
   index: 2
 - name: Australian Institute of Marine Science, Townsville, Qld 4810, Australia
   index: 3

@@

The Australian Institute of Marine Science (AIMS) has a long tradition in
measuring and monitoring a series of environmental parameters along the
tropical coast of Australia. These parameters include long-term record of sea
surface temperature, wind characteristics, atmospheric temperature, pressure,
chlorophyll-a data, among many others. The AIMS Data Centre team has recently
developed the [AIMS Data Platform API][1] which is a *REST API* providing
JSON-formatted data to users. `dataaimsr` is an **R package** written to
allow users to communicate with the AIMS Data Platform API using an API key
and a few convenience functions to interrogate and understand the datasets
that are available to download.

@@

Before loading the package, a user needs to download and store their personal
[AIMS Data Platform API Key][4]---we recommend it storing it in `.Renviron`.

@@

@@

The first step would be to visualise the dataset. We do this by
mapping all available sites. First we need to store the DOI for the
target dataset---this is done via the function `aims_data_doi`. We
can then download the summary information for the Sea Water Temperature 
Loggers dataset using the main function called `aims_data`:

```{r, message = FALSE, warning = FALSE}
library(dataaimsr)
# see ?aims_data_doi for argument names
ssts_doi <- aims_data_doi("temp_loggers")
sdata <- aims_data(ssts_doi, api_key = NULL,
                   summary = "summary-by-series")
head(sdata)
```

Setting the argument `api_key = NULL` means that `dataaimsr` will
automatically search for the user's API key stored in `.Renviron`.
The `summary` argument here is key. It should only be flagged when the
user wants an overview of the available data. Again, this currently
implemented for the Sea Water Temperature Loggers dataset. One can
visualise `summary-by-series` or `summary-by-deployment`. The output of
`aims_data` when summary is `NA` (the default) is a `data.frame`.

@@

Also note that there are three columns containing the total number of 
observations in a series: `uncal_obs`, `cal_obs` and `qc_obs`, which 
respectively stand for uncalibrated, calibrated, and quality-controlled 
observations. **Mark, we need a few sentences explaining the quality control**
**procdure**. One can visualise this data, for instance, by plotting them
on a map of Australia, while colouring based on the total amount of calibrated
observations (Fig. \@ref(fig:summary)).

```{r echo = FALSE, eval = FALSE}
library(tidyverse)
library(ggmap)
sdata <- sdata %>%
  dplyr::mutate(cols = cal_obs * 1e-3) %>%
  tidyr::drop_na(lon, lat)
lab <- expression(atop("# Calibrated obs.",
                       paste("(" %*% 10^3, ")")))
ozmap <- ggmap::get_stamenmap(bbox = c(left = 94, bottom = -45,
                                       right = 162, top = 0),
                              zoom = 4, maptype = "watercolor")
out <- ggmap::ggmap(ozmap) +
  theme_void() +
  geom_point(data = sdata,
             mapping = aes(x = lon, y = lat,
                           colour = cols,
                           fill = cols),
             shape = 16, alpha = 0.8) +
  scale_colour_gradient(name = lab,
                        low = "dodgerblue4",
                        high = "darkred") +
  scale_fill_gradient(name = lab,
                      low = "dodgerblue4",
                      high = "darkred")
ggsave("paper/summary.png", out, width = 7, height = 4.36, units = "in",
       device = "png", dpi = 300)
```

@@

In the case of the Weather Station dataset, the user can call a
the `filter_values` function which allows one to query what
sites, series and parameters are available for both datasets:

```{r, message = FALSE, warning = FALSE}
weather_doi <- aims_data_doi("weather")
filter_values(weather_doi, filter_name = "series") %>%
  head()
```

@@

Filters are the last important information the user needs to know to 
master the navigation and download of AIMS monitoring datasets. Each 
dataset can filtered by attributes which can be exposed with the function `expose_attributes`:

```{r, message = FALSE, warning = FALSE}
expose_attributes(weather_doi)
expose_attributes(ssts_doi)
```

The help file (see `?expose_attributes`) contains the details about what
each filter targets. So, having an understanding of the summaries and what
filters are available provide the user with a great head start.

@@

```{r, message = FALSE, warning = FALSE}
wdata_a <- aims_data(weather_doi, api_key = NULL,
                     filters = list(site = "Yongala",
                                    from_date = "2018-01-01",
                                    thru_date = "2018-01-02"))
```

The output of `aims_data` when summary is omitted (the default) is a list
containing three elements:

@@

- `data` an output `data.frame`

```{r}
wdata_a$metadata
```

```{r}
wdata_a$citation
```

```{r}
head(wdata_a$data)
```

Note that there are numerous parameters available for this site at the
specified time:

```{r}
unique(wdata_a$data$parameter)
```

And the actual measurements are either raw or quality-controlled (Fig. \@ref(fig:wind)).

```{r echo = FALSE, eval = FALSE}
out <- wdata_a$data %>%
  dplyr::filter(grepl("Wind Speed", parameter)) %>%
  ggplot(data = .) +
    geom_line(aes(x = time, y = qc_val, colour = parameter)) +
    labs(x = "Date",
         y = "Wind speed (km / h)",
         colour = "Site",
         title = "AIMS Weather stations",
         subtitle = "Yongala (2018)") +
    theme_bw()
ggsave("paper/wind.png", out, width = 7.4, height = 5.2, units = "in",
       device = "png", dpi = 300)
```

(ref:fig-wind) Yongala wreck wind speed profiles between the first and second of January 2018.

@@

```{r, message = FALSE, warning = FALSE}
wdata_b <- aims_data(weather_doi,
                     api_key = NULL,
                     filters = list(series_id = 64,
                                    from_date = "1991-10-18T06:00:00",
                                    thru_date = "1991-10-18T12:00:00"))$data
range(wdata_b$time)
```

@@

# Acknowledgements

The development of `dataimsr` was supported by ... Names to be added to the list.

>>>>

---
title: 'dataaimsr: An R Client for the Australian Institute of Marine Science Data Platform API which provides easy access to AIMS Data Platform'
tags:
  - R
  - sea surface temperature
  - weather stations
  - long-term environmental monitoring
  - Australia
  - API
authors:
  - name: Diego R. Barneche
    orcid: 0000-0002-4568-2362
    affiliation: "1, 2"
  - name: Greg Coleman
    affiliation: "3"
  - name: Duncan Fermor
    affiliation: "3"
  - name: Eduardo Klein
    affiliation: "3"
  - name: Tobias Robinson
    affiliation: "3"
  - name: Jason Smith
    affiliation: "3"
  - name: Jeffrey L. Sheehan
    affiliation: "3"
  - name: Shannon Dowley
    affiliation: "3"
  - name: Dean Ditton
    affiliation: "3"
  - name: Kevin Gunn
    affiliation: "3"
  - name: Gavin Ericson
    affiliation: "3"
  - name: Murray Logan
    affiliation: "3"
  - name: Mark Rehbein
    affiliation: "3"
affiliations:
 - name: Australian Institute of Marine Science, Crawley, WA 6009, Australia
   index: 1
 - name: The Indian Ocean Marine Research Centre, The University of Western Australia, Crawley, WA 6009, Australia
   index: 2
 - name: Australian Institute of Marine Science, Townsville, Qld 4810, Australia
   index: 3

@@

The Australian Institute of Marine Science (AIMS) has a long tradition in
measuring and monitoring a series of environmental parameters along the
tropical coast of Australia. These parameters include long-term record of sea
surface temperature, wind characteristics, atmospheric temperature, pressure,
chlorophyll-a data, among many others. The AIMS Data Centre team has recently
developed the [AIMS Data Platform API][1] which is a *REST API* providing
JSON-formatted data to users. `dataaimsr` is an **R package** written to
allow users to communicate with the AIMS Data Platform API using an API key
and a few convenience functions to interrogate and understand the datasets
that are available to download. In doing so, it allows the user to
fully explore these datasets in R in whichever capacity they want (e.g.
data visualisation, statistical analyses, etc). The package itself contains
a `plot` method which allows the user to plot summaries of the different types
of dataset made available by the API.

@@

Before loading the package, a user needs to download and store their personal
[AIMS Data Platform API Key][4]---we strongly encourage users to
maintain their API key as a private, locally hidden environment variable
(`AIMS_DATAPLATFORM_API_KEY`) in the `.Renviron` file for
automatic loading at the start of an R session.

@@

`dataaimsr` imports the packages *httr* [@httrcit], *jsonlite* [@jsonlitecit],
*parsedate* [@parsedatecit], *dplyr* [@dplyrcit], *tidyr* [@tidyrcit],
*rnaturalearth* [@rnaturalearthcit], *sf* [@sfcit], *ggplot2* [@ggplot2cit],
*ggrepel* [@ggrepelcit] and *curl* [@curlcit].

@@

The first step would be to visualise the dataset. We do this by
mapping all available sites. For example, we download the summary information
for the Sea Water Temperature Loggers dataset using the main function called
`aims_data`:

```{r, message = FALSE, warning = FALSE}
library(dataaimsr)
sdata <- aims_data("temp_loggers", api_key = NULL,
                   summary = "summary-by-series")
head(sdata)
```

Setting the argument `api_key = NULL` means that `dataaimsr` will
automatically search for the user's API key stored in `.Renviron`.
The `summary` argument here is key. It should only be flagged when the
user wants an overview of the available data. Again, this currently
implemented for the Sea Water Temperature Loggers dataset. One can
visualise `summary-by-series` or `summary-by-deployment`. The output of
`aims_data` is a `data.frame` of class `aimsdf`.

@@

Also note that there are three columns containing the total number of 
observations in a series: `uncal_obs`, `cal_obs` and `qc_obs`, which 
respectively stand for uncalibrated, calibrated, and quality-controlled 
observations. Calibrated and quality-controlled are generally the same.
Instruments are routinely calibrated (mostly once a year) in a temperature-controlled water bath and corrections applied to the data. After
calibration, all data records are quality controlled based on the following
tests: 1) clip to in-water only data, using deployment's metadata, 2)
impossible value check: data outside a fixed temperature range (14˚C – 40˚C)
is flagged as bad data, 3) spike test: individual extreme values are flagged
as probably bad according to the algorithm presented in @morelo2014methods and
4) Excessive gradient test: pairs of data that present a sudden change in the
slope are flagged as probably bad [@toma2016acta]. If any data record fails at
least one of the tests, a QC flag equal to 2 is returned, otherwise, the QC
flag is set to 1.

One can visualise this data, for instance, by plotting them
on a map of Australia, while colouring based on the total amount of calibrated
observations (Fig. \@ref(fig:summary)).

`aimsdf` objects can be plotted using the `plot` function. For summary data
such as `sdata`, plot will always generate a map with the points around
Australia and associated regions, coloured by the number of calibrated
observations:

```{r echo = FALSE, eval = FALSE}
out <- plot(sdata, ptype = "map")
ggsave("paper/summary.png", out, width = 7, height = 4.36, units = "in",
       device = "png", dpi = 300)
```

@@

In the case of the Weather Station dataset, the user can call a
the `aims_filter_values` function which allows one to query what
sites, series and parameters are available for both datasets:

```{r, message = FALSE, warning = FALSE}
aims_filter_values("weather", filter_name = "series") %>%
  head()
```

@@

Filters are the last important information the user needs to know to 
master the navigation and download of AIMS monitoring datasets. Each 
dataset can filtered by attributes which can be exposed with the function
`aims_expose_attributes`:

```{r, message = FALSE, warning = FALSE}
aims_expose_attributes("weather")
aims_expose_attributes("temp_loggers")
```

The help file (see `?aims_expose_attributes`) contains the details about what
each filter targets. So, having an understanding of the summaries and what
filters are available provide the user with a great head start.

@@

```{r, message = FALSE, warning = FALSE}
wdata_a <- aims_data("weather", api_key = NULL,
                     filters = list(site = "Yongala",
                                    from_date = "2018-01-01",
                                    thru_date = "2018-01-02"))
```

The returned `aimsdf` object in this case has attributes which give us
summary crucial information:

@@

- `parameters` an output `data.frame`

These can be directly extracted using the convenience functions
`aims_metadata`, `aims_citation` and `aims_parameters`, e.g.:

```{r}
aims_metadata(wdata_a)
```

This example data contains multiple parameters available for this site at the
specified time, and the actual measurements are either raw or
quality-controlled. For monitoring data (i.e. when `summary = NA` in a
`aims_data` call), we can either visualise the data as a time series broken
down by parameter, or a map showing the sites with some summary info. If the
parameters are not specified, then `dataaimsr` will plot a maximum of 4
parameters chosen at random for a time series plot. Alternatively the user can
specify which parameters are to be plotted (Fig. \@ref(fig:wind)).

```{r echo = FALSE, eval = FALSE}
# check parameters with aims_parameters(wdata_a)
out <- plot(wdata_a, ptype = "time_series",
            pars = c("Water Pressure", "Chlorophyll"))
ggsave("paper/wind.png", out, width = 8.5, height = 3.9, units = "in",
       device = "png", dpi = 300)
```

(ref:fig-wind) Yongala wreck profiles for water pressure and chlorophyll-a between the first and second of January 2018.

@@

```{r, message = FALSE, warning = FALSE}
wdata_b <- aims_data("weather", api_key = NULL,
                     filters = list(series_id = 64,
                                    from_date = "1991-10-18T06:00:00",
                                    thru_date = "1991-10-18T12:00:00"))
range(wdata_b$time)
```

### Methods

Objects of class `aimsdf` have associated `plot`, `print` and `summary`
methods.

### Data citation

Whenever using `dataaimsr`, we ask the user to not only cite this paper, but
also any data used in an eventual publication. Citation data can be extracted
from a dataset using the function `aims_citation`:

```{r, message = FALSE, warning = FALSE}
aims_citation(wdata_b)
```

## Sister web tool

The Time Series Explorer (https://apps.aims.gov.au/ts-explorer/) is an
interactive web-based application that visualises large time series datasets.
The application utilises the AIMS Data Platform API to dynamically query data
according to user selection and visualise the data as line graphs. Series are
able to be compared visually. For large series, data are aggregated to daily
averages and displayed as minimum, maximum and mean. When the user 'zooms in'
sufficiently, the data will be displayed as non-aggregate values
(Fig. \@ref(fig:tssa)). This technique is being used to ensure the application
performs well with large time series.

(ref:fig-tssa) Interactive discovery and visualisation of data series.

```{r tssa, echo = FALSE, out.width = "100%", fig.cap="(ref:fig-tssa)", cache = TRUE}
knitr::include_graphics("tssa.png")
```

The user can then download the displayed data as CSV or obtain a R code
snippet that shows how to obtain the data using the dataaimsr package
(Fig. \@ref(fig:tssb)). In this way, a user can easily explore and discover
datasets and then quickly and easily have this data in their R environment for
additional analysis.

(ref:fig-tssb) Download/Export displayed data via R snippet.

```{r tssb, echo = FALSE, out.width = "100%", fig.cap="(ref:fig-tssb)", cache = TRUE}
knitr::include_graphics("tssb.png")
```

@@

The API is still a work in progress. We are working on ways to better
facilitate data visualisation and retrieval, and also we are trying to
standardise the outputs from the different datasets as much as possible. In the
future, we envision that `dataaimsr` will also provide access to other
monitoring datasets collected by AIMS.
